---
title: "AL selected human sequences expression analysis Report"
author: "Emmanuel Cazottes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,          # Show code chunks in output
  warning = FALSE,      # Don't show warnings
  message = FALSE,      # Don't show messages
  fig.width = 10,       # Default figure width
  fig.height = 6,       # Default figure height
  fig.align = "center"  # Center align figures
)

# Load commonly used packages
library(stringr)
library(tidyverse)
library(reshape2)
library(ggpubr)
library(ggsci)
library(gridExtra)

# Custom ggplot theme settings
publication_theme <- theme_pubr() +
  theme(
    # Text elements
    axis.text = element_text(size = 7, color = "black"),
    axis.title = element_text(size = 10, face = "plain"),
    plot.title = element_text(size = 10, face = "plain", hjust = 0.5),
    legend.title = element_text(size = 10, face = "plain"),
    legend.text = element_text(size = 7),
    
    # Plot elements
    axis.line = element_line(size = 0.5, color = "black"),
    axis.ticks = element_line(size = 0.5, color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    
    # Legend positioning and appearance
    legend.position = "right",
    legend.box.background = element_rect(color = "white"),
    legend.background = element_rect(fill = "white", color = NA),
    
    # Panel appearance
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white")
  )

theme_set(publication_theme)
```

This notebook analyze the expression levels of the most and least selected human K562 sequences by AL methods (top/bottom 10%) compared to random selection and the AL pool. K562 MPRA data from Gosai et al 2024 are used to annotate expression of selected sequences.

Only the K562 data was used to train the model so the HepG2 and SKNHSH columns are removed

# Data import and preprocessing

Sequences expression as measured in K562 MPRA

```{r}
mpra = read.table("reference/41586_2024_8070_MOESM4_ESM.txt", header = T, sep = "\t")
mpra = mpra[,-c(2,3,4,5,length(mpra))] #Remove the columns I won't use
mpra = mpra[,c(1,2,5)] #Keeping the K562 related columns
```

Load the sequence data

```{r}
load.sequences = function(seq.tab){
  seq = read.table(seq.tab, header = F, sep = "\t", col.names = c("IDs", "sequence","method"))
  seq = seq[,-2]
  return(seq)
}

alPool = load.sequences("sequences/pool.tsv")
alPool$rounds = "AL pool"
alPool$method = "AL pool"
alPool$al.method = "AL pool"

tnb = load.sequences("sequences/human_top_n_bottom_10.tsv")
tnb[grep("1x60", tnb$method),"rounds"] = "1x60k"
tnb[grep("1x60", tnb$method,invert = T),"rounds"] = "3x20k"
tnb[grep("uncertainty_((1x60)|(3x20))_bottom.*", tnb$method),"al.method"] = "Least uncertain"
tnb[grep("uncertainty_((1x60)|(3x20))_top.*", tnb$method),"al.method"] = "Most uncertain"
tnb[grep("diversity_((1x60)|(3x20))_bottom.*", tnb$method),"al.method"] = "Least diverse"
tnb[grep("diversity_((1x60)|(3x20))_top.*", tnb$method),"al.method"] = "Most diverse"
```



# Merge sequence ID with expression values

Merge the mpra data with the sequence data

```{r}
tnb.mpra = merge(tnb, mpra,by="IDs")
alPool.mpra = merge(alPool, mpra, by="IDs")

#Random selection
rnd.mpra = alPool.mpra[sample(seq(1,nrow(alPool.mpra),1),100000),]
rnd.mpra$method = "Random selection"
rnd.mpra$rounds = "Random selection"
rnd.mpra$al.method = "Random selection"

```

Bind everything together for plotting

```{r}
tnb.mpra.3x20 = subset(tnb.mpra, rounds=="3x20k")
tnb.mpra.1x60 = subset(tnb.mpra, rounds=="1x60k")
```

```{r}
seq.mpra.3x20 = rbind( tnb.mpra.3x20, rnd.mpra, alPool.mpra)
seq.mpra.1x60 = rbind(tnb.mpra.1x60, rnd.mpra, alPool.mpra)

#Levels for plotting
seq.mpra.3x20$al.method = factor(seq.mpra.3x20$al.method, levels = c("AL pool", "Random selection",  "Least uncertain","Most uncertain","Least diverse","Most diverse"))
seq.mpra.1x60$al.method = factor(seq.mpra.1x60$al.method, levels = c("AL pool", "Random selection",  "Least uncertain","Most uncertain","Least diverse","Most diverse"))
```

Test statistical differences between the groups

```{r}
# compute the Mann-Whitney pvalue
mannWhitneyTest = function(data){
  df.pval = map_dfr(levels(data$al.method), function(level){
  group1 = subset(data, al.method==level)$K562_log2FC
  group2 = subset(data, al.method=="AL pool")$K562_log2FC
  
  mwt = wilcox.test(group1, group2)
  
  data.frame(al.method=level,
             pvalue=signif(mwt$p.value, digits = 2),
             statistic=mwt$statistic)
  })
  
  df.pval$pvalue = ifelse((df.pval$pvalue==0 | df.pval$pvalue<=2.2e-16),"p < 2.2e-16", paste0("p = ",df.pval$pvalue)) #R reports 0 because the pvalue is smaller than R's standard floating-point precision

  return(df.pval)
  }
```

```{r}
pval60k = mannWhitneyTest(seq.mpra.1x60)
pval60k$y.axis = 10
pval3x20 = mannWhitneyTest(seq.mpra.3x20)
pval3x20$y.axis = 10
```

# Plots

```{r}
boxplot.fc.3x20 = ggplot(seq.mpra.3x20 , aes(x=al.method, y=K562_log2FC))+
  geom_violin(width=1.25, adjust=3)+  
  geom_boxplot(width=0.1,outliers=F)+
    geom_text(data = pval3x20, aes(x=al.method,y=y.axis, label = pvalue),size = 3)+
  xlab("")+
  ylab("log2(FC K562)")+
  ggtitle("K562 - 3x20")+
    theme(aspect.ratio = c(0.85/1),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10))

boxplot.fc.1x60 = ggplot(seq.mpra.1x60 , aes(x=al.method, y=K562_log2FC))+
  geom_violin(width=1.25, adjust=3)+  
  geom_boxplot(width=0.1,outliers=F)+
    geom_text(data = pval60k, aes(x=al.method,y=y.axis, label = pvalue),size = 3)+
  xlab("")+
  ylab("log2(FC K562)")+
    ggtitle("K562 - 1x60")+
    theme(aspect.ratio = c(0.85/1),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10))

```
```{r selected-sequences-plots, fig.cap="Selected sequences expression analysis"}
print(boxplot.fc.3x20)
print(boxplot.fc.1x60)
```

