---
title: "Human TFBS Analysis Report"
author: "Emmanuel Cazottes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,          # Show code chunks in output
  warning = FALSE,      # Don't show warnings
  message = FALSE,      # Don't show messages
  fig.width = 10,       # Default figure width
  fig.height = 6,       # Default figure height
  fig.align = "center"  # Center align figures
)

# Load commonly used packages
library(stringr)
library(tidyverse)
library(reshape2)
library(ggpubr)
library(ggsci)
library(gridExtra)
library(effectsize)

# Custom ggplot theme settings
publication_theme <- theme_pubr() +
  theme(
    # Text elements
    axis.text = element_text(size = 7, color = "black"),
    axis.title = element_text(size = 10, face = "plain"),
    plot.title = element_text(size = 10, face = "plain", hjust = 0.5),
    legend.title = element_text(size = 10, face = "plain"),
    legend.text = element_text(size = 7),
    
    # Plot elements
    axis.line = element_line(size = 0.5, color = "black"),
    axis.ticks = element_line(size = 0.5, color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    
    # Legend positioning and appearance
    legend.position = "right",
    legend.box.background = element_rect(color = "white"),
    legend.background = element_rect(fill = "white", color = NA),
    
    # Panel appearance
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white")
  )

theme_set(publication_theme)
```

Analyse the TFBS content of human selected sequences, AL pool sequences, Top/Bottom 10% selected sequences and randomly selected sequences. 

This notebook take as input the `fimo.tsv` file obtained by doing motif scanning with FIMO with the HOCOMOCO 13 as TF motifs database and thresdholding on FDR 0.01. The custom background passed to FIMO, was generated using nucleotide frequencies derived from the AL pool sequences using the `fasta-get-markov` tool from the MEME suite. 

Filtering for K562 TF using the GTEX (V8) Whole Blood annotations. HOCOMOCO13 is using Uniprot IDs while GTEX uses gene symbols, the `tf_masterlist.tsv` file was downloaded from the Autosome.org website and contains the TF Uniprot IDs and their corresponding gene symbols.


# Data import and preprocessing

Selection of TFs expressed in K562

```{r}
#Select TF expressed in K562
gtex = read.table("reference/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz", header = T, sep = "\t",skip = 2)

#Trim the table to remove the mouse orthologs 
tf.masterlist = read.table("reference/tf_masterlist.tsv", header = T, sep = "\t", fill = T)
tf.masterlist = tf.masterlist[grep("HUMAN", tf.masterlist$curated.uniprot_id),]
tf.masterlist$curated.uniprot_id = str_split(tf.masterlist$curated.uniprot_id, "_",simplify = T)[,1]
tf.masterlist = tf.masterlist[,c("curated.uniprot_id","auto.gene_symbol")]

#Select for TFs expressed in WB
#Merging with the TF masterlist from the autosome team is important because the TFs IDs are from uniprot and not gene symbols
gtex = merge(gtex[,c("Description","Whole.Blood")], tf.masterlist, by.x="Description", by.y="auto.gene_symbol")
gtex.selected.TF = subset(gtex, (Whole.Blood>=1))$curated.uniprot_id

```

Function to import the fimo.tsv file containing the TFBS for each selected sequences and the original sequence file to recover the sequences without TFBS.

```{r}
preprocess.tf = function(data.tf, path.seq){
  if (is.data.frame(data.tf)) {
      seq = read.table(path.seq, header = F, sep = "\t", col.names = c("sequence_name", "sequence", "method"))
      seq = seq[,-2]
      tf.motifs = merge(seq,data.tf,by = "sequence_name",all.x = T)
      tf.motifs = tf.motifs[,-3] # REmove the duplicated method column
      colnames(tf.motifs) = c("sequence_name", "method", "motif_id")
      tf.motifs$motif_id = ifelse(is.na(tf.motifs$motif_id),0,tf.motifs$motif_id)
      return(tf.motifs)
  } else if (file.exists(data.tf)){
    tf.motifs = read.table(data.tf, header = T, sep = "\t")
    tf.motifs = tf.motifs[,-c(2,4,5,6,7,8,9,10)]
    tf.motifs$motif_id = unlist(lapply(str_split(tf.motifs$motif_id, "\\."), function(x){x[1]})) # Get the TF name
    tf.motifs$method = unlist(lapply(str_split(tf.motifs$sequence_name, "\\|"), function(x){x[2]})) # Get the method name
    tf.motifs$sequence_name = unlist(lapply(str_split(tf.motifs$sequence_name, "\\|"), function(x){x[1]})) # Get the SeqID
    
    tf.motifs = subset(tf.motifs, motif_id %in% gtex.selected.TF) #Keep only K562 TFs
    #Train pool sequences
    seq = read.table(path.seq, header = F, sep = "\t", col.names = c("sequence_name", "sequence", "method"))
    seq = seq[,-2]

    tf.motifs = merge(seq,tf.motifs,by = "sequence_name",all.x = T)
    tf.motifs = tf.motifs[,-3] # REmove the duplicated method column
    colnames(tf.motifs) = c("sequence_name", "method", "motif_id")
    tf.motifs$motif_id = ifelse(is.na(tf.motifs$motif_id),0,tf.motifs$motif_id)
    return(tf.motifs)
  }
}
```

```{r}
#AL pool
tf.motifs.al = preprocess.tf("sequence_analysis/hocomoco_ALpool_FDR0.01/fimo.tsv", "sequences/pool.tsv")

#Selected sequences
tf.motifs.selected = preprocess.tf(tf.motifs.al, "sequences/master_output.tsv")
#Remove Seed information
tf.motifs.selected$method = str_remove(tf.motifs.selected$method, "(_[1-5])")
tf.motifs.selected$method = gsub("(_1x60)", "", tf.motifs.selected$method)

#Top/Bottom 10% selected sequences
tf.motifs.tnb = preprocess.tf(tf.motifs.al, "sequences/human_top_n_bottom_10.tsv")
tf.motifs.tnb$method = gsub("1x60","60k",tf.motifs.tnb$method) #homogenous naming

```

Some data are duplicated across pairs : RNN+CNN and All arch, let's keep them only for the RNN

```{r}
tf.motifs.selected = subset(tf.motifs.selected, 
                                  !(method %in% c("rnn-cnn_attn","diff_arch_attn","rnn-cnn_60k_attn","diff_arch_60k_attn","rnn-cnn_cnn","diff_arch_cnn","rnn-cnn_60k_cnn","diff_arch_60k_cnn")))

```

```{r}
# Format the labels for selected sequences
no.seed = str_remove(tf.motifs.selected$method, "(_[1-5]_1x60)|(_[1-5]_3x20)")
tf.motifs.selected$method = no.seed
```


# Count the total number of TFBS per sequence

Function to count the number of motifs for each TF per sequence and then the total number of motifs per sequences

```{r}
counts.tf = function(tf.df){
  #Counts per TF of each sequence and method
  counts.per.tf = subset(tf.df, motif_id!=0) %>%
  group_by(sequence_name,method, motif_id) %>%
  summarise(counts = n())

#Add the sequences with no TF motif found
  TF.0.counts = subset(tf.df, motif_id==0)
  TF.0.counts$counts = 0 

  counts.per.tf = rbind(counts.per.tf, TF.0.counts)
#Counts of TF per sequences
  tf.per.seq = counts.per.tf %>%
    group_by(sequence_name,method) %>%
    summarise(counts = sum(counts))
  
  return(tf.per.seq)
}
```

```{r}
count.selected = counts.tf(tf.motifs.selected)
counts.tnb = counts.tf(tf.motifs.tnb)
counts.al = counts.tf(tf.motifs.al)
```

Randomly selected sequences from the AL pool to use as a baseline for the Cliff's delta. 15 random sampling of 20 000 sequences.

```{r}
random.sampling = function(df, nb.rnd, nb.sample){
  value.list = list()
  size = seq(1, nrow(df), by=1)
  i= 1
  while (i <= nb.sample) {
    rnd.size = sample(size, nb.rnd)
    value.list[[i]] = df[rnd.size,]
    value.list[[i]]$method = paste0("RandomSampling_",i)
    i = i +1
  }
  return(value.list)
}

counts.tf.rnd = random.sampling(counts.al,20000,15 )
```

# Cliff's delta

Compute the cliff's delta on the number of TFBS per sequence for each architecture/model pair and the AL pool. The question being do we have a different TFBS motif content (more or less TF)

## Selected sequences

```{r}
count.selected.list = count.selected %>% 
  ungroup() %>% #Important to ungroup because the count function is grouping on sequence names
  group_split(method)
```

Perform Cliff's delta between the TFBS content of selected sequences to the AL pool.
```{r}
cliff.d.selected = lapply(count.selected.list, function(x){
  cliffs_delta(x$counts, counts.al$counts)
})

cliff.d.selected = do.call("rbind", cliff.d.selected)

cliff.d.selected$method = unique(count.selected$method)

cliff.d.selected$model.arch = unlist(lapply(str_split(cliff.d.selected$method, "_"), function(x){x[length(x)]}))
cliff.d.selected$al.method = unlist(lapply(str_split(cliff.d.selected$method, "_(attn|rnn|cnn)"), function(x){x[1]}))
cliff.d.selected$model.arch = toupper(cliff.d.selected$model.arch)
```

Splitting the data between the 3x20 and 1x60 configurations for plotting.

```{r}
cliff.d.selected.60k = cliff.d.selected[grep(pattern = "60k", cliff.d.selected$method),]
cliff.d.selected.60k$model.arch = unlist(lapply(str_split(cliff.d.selected.60k$method, "_"), function(x){x[length(x)]}))
cliff.d.selected.60k$al.method = unlist(lapply(str_split(cliff.d.selected.60k$method, "_(attn|rnn|cnn)"), function(x){x[1]}))
cliff.d.selected.60k$al.method = gsub("_60k", "" ,cliff.d.selected.60k$al.method)
cliff.d.selected.60k$model.arch = toupper(cliff.d.selected.60k$model.arch)
cliff.d.selected.60k$model.arch = factor(cliff.d.selected.60k$model.arch, levels = c("RNN","CNN","ATTN"))
```

```{r}
cliff.d.selected.3x20k = cliff.d.selected[grep(pattern = "60k", cliff.d.selected$method, invert = T),]
cliff.d.selected.3x20k$method = gsub("_3x20", "" ,cliff.d.selected.3x20k$method)
cliff.d.selected.3x20k$model.arch = unlist(lapply(str_split(cliff.d.selected.3x20k$method, "_"), function(x){x[length(x)]}))
cliff.d.selected.3x20k$al.method = unlist(lapply(str_split(cliff.d.selected.3x20k$method, "_(attn|rnn|cnn)"), function(x){x[1]}))

cliff.d.selected.3x20k$model.arch = toupper(cliff.d.selected.3x20k$model.arch)
cliff.d.selected.3x20k$model.arch = factor(cliff.d.selected.3x20k$model.arch, levels = c("RNN","CNN","ATTN"))
```

## Cliff's delta for randomly selected sequences

```{r}
cliff.d.rnd = lapply(counts.tf.rnd, function(x){
  cliffs_delta(x$counts, counts.al$counts)
})

cliff.d.rnd = do.call("rbind", cliff.d.rnd)

cliff.d.rnd = data.frame(
  r_rank_biserial = mean(cliff.d.rnd$r_rank_biserial),
  CI=0.95,
  CI_low = min(cliff.d.rnd$r_rank_biserial),
  CI_high = max(cliff.d.rnd$r_rank_biserial),
  al.method="Random selection"
)
```

## Control sequences

Cliff's delta for control sequences : AL pool and Top/Bottom 10% most selected sequences

```{r}
counts.controlSeq = rbind(counts.al, counts.tnb)
counts.controlSeq = counts.controlSeq %>%
  group_by(method) %>%
  group_split()
```


```{r}
cliff.d.controls = lapply(counts.controlSeq, function(x){
  cliffs_delta(x$counts, counts.al$counts)
})

cliff.d.controls = do.call("rbind", cliff.d.controls)
cliff.d.controls$al.method = unlist(lapply(counts.controlSeq, function(x){unique(x$method)}))
cliff.d.controls$al.method = gsub("1x60", "60k", cliff.d.controls$al.method)

cliff.d.controls = rbind(cliff.d.controls, cliff.d.rnd)
```

```{r}
cliff.d.controls.3x20k = cliff.d.controls[grep(pattern = "60k", cliff.d.controls$al.method,invert = T),]
#Rename labels for plotting
cliff.d.controls.3x20k$al.method = gsub("diversity_3x20_bottom10", "Least diverse", cliff.d.controls.3x20k$al.method)
cliff.d.controls.3x20k$al.method = gsub("diversity_3x20_top1010", "Most diverse", cliff.d.controls.3x20k$al.method)
cliff.d.controls.3x20k$al.method = gsub("uncertainty_3x20_bottom10", "Least uncertain", cliff.d.controls.3x20k$al.method)
cliff.d.controls.3x20k$al.method = gsub("uncertainty_3x20_top1010", "Most uncertain", cliff.d.controls.3x20k$al.method)
#Order the labels
cliff.d.controls.3x20k$al.method = factor(cliff.d.controls.3x20k$al.method, 
                                levels = c("ALpool", "Random selection", "Least uncertain", "Most uncertain", "Least diverse", "Most diverse"))
```

```{r}
cliff.d.controls.60k = cliff.d.controls[grep(pattern = "(60k)|(ALpool)|(Random selection)", cliff.d.controls$al.method),]
cliff.d.controls.60k$al.method = gsub("diversity_60k_bottom10", "Least diverse", cliff.d.controls.60k$al.method)
cliff.d.controls.60k$al.method = gsub("diversity_60k_top1010", "Most diverse", cliff.d.controls.60k$al.method)
cliff.d.controls.60k$al.method = gsub("uncertainty_60k_bottom10", "Least uncertain", cliff.d.controls.60k$al.method)
cliff.d.controls.60k$al.method = gsub("uncertainty_60k_top1010", "Most uncertain", cliff.d.controls.60k$al.method)
#Order the labels
cliff.d.controls.60k$al.method = factor(cliff.d.controls.60k$al.method, 
                                levels = c("ALpool", "Random selection", "Least uncertain", "Most uncertain", "Least diverse", "Most diverse"))
```

# Plots

## Selected sequences 

```{r}
cliff.d.selected.3x20.plot = ggplot(cliff.d.selected.3x20k, aes(al.method, y=r_rank_biserial))+
  geom_point(size=2)+
  geom_linerange(aes(ymin = CI_low, ymax = CI_high), width = 0.5, colour="black")+
  geom_hline(yintercept = cliff.d.rnd$r_rank_biserial,linetype=2)+
  scale_y_continuous(limits = c(-0.3, 0.3), breaks = round(seq(-0.3, 0.3, by = 0.1),2),expand = expansion(mult = 0.05))+
  facet_wrap(~model.arch, scales = "free_x")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),axis.text = element_text(size=7),plot.margin = margin(5, 5, 5, 5, "mm"))+
  labs(x="", y="Cliff's delta", title = "Human - 3x20")

cliff.d.selected.1x60.plot = ggplot(cliff.d.selected.60k, aes(al.method, y=r_rank_biserial))+
  geom_point(size=2)+
  geom_linerange(aes(ymin = CI_low, ymax = CI_high), width = 0.5, colour="black")+
  geom_hline(yintercept = cliff.d.rnd$r_rank_biserial,linetype=2)+
  scale_y_continuous(limits = c(-0.3, 0.3), breaks = round(seq(-0.3, 0.3, by = 0.1),2),expand = expansion(mult = 0.05))+
  facet_wrap(~model.arch, scales = "free_x")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),axis.text = element_text(size=7),plot.margin = margin(5, 5, 5, 5, "mm"))+
  labs(x="", y="Cliff's delta", title = "Human - 1x60")
```

```{r selected-sequences-plots, fig.cap="Selected sequences TFBS analysis"}
print(cliff.d.selected.3x20.plot)
print(cliff.d.selected.1x60.plot)
```

## Control sequences

```{r}
cliff.d.ControlSeq.3x20.plot = ggplot(cliff.d.controls.3x20k, aes(al.method, y=r_rank_biserial))+
  geom_point(size=2)+
  geom_linerange(aes(ymin = CI_low, ymax = CI_high), width = 0.5, colour="black")+
  geom_hline(yintercept = cliff.d.rnd$r_rank_biserial,linetype=2)+
  scale_y_continuous(limits = c(-0.4, 0.4), breaks = round(seq(-0.4, 0.4, by = 0.1),2),expand = expansion(mult = 0.05))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text = element_text(size=7),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        aspect.ratio = c(1/0.75))+
  labs(x="", y="Cliff's delta", title = "Human - 3x20")

cliff.d.ControlSeq.1x60.plot = ggplot(cliff.d.controls.60k, aes(al.method, y=r_rank_biserial))+
  geom_point(size=2)+
  geom_linerange(aes(ymin = CI_low, ymax = CI_high), width = 0.5, colour="black")+
  geom_hline(yintercept = cliff.d.rnd$r_rank_biserial,linetype=2)+
  scale_y_continuous(limits = c(-0.4, 0.4), breaks = round(seq(-0.4, 0.4, by = 0.1),2),expand = expansion(mult = 0.05))+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text = element_text(size=7),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        aspect.ratio = c(1/0.75))+
  labs(x="", y="Cliff's delta", title = "Human - 1x60")
```

```{r Control-sequences-plots, fig.cap="Cliff's delta for control sequences"}
print(cliff.d.ControlSeq.3x20.plot)
print(cliff.d.ControlSeq.1x60.plot)
```