---
title: "Analysis Report"
author: "Emmanuel Cazottes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,          # Show code chunks in output
  warning = FALSE,      # Don't show warnings
  message = FALSE,      # Don't show messages
  fig.width = 10,       # Default figure width
  fig.height = 6,       # Default figure height
  fig.align = "center"  # Center align figures
)

# Load commonly used packages
library(stringr)
library(tidyverse)
library(reshape2)
library(ggpubr)
library(ggsci)
library(gridExtra)

# Custom ggplot theme settings
publication_theme <- theme_pubr() +
  theme(
    # Text elements
    axis.text = element_text(size = 7, color = "black"),
    axis.title = element_text(size = 10, face = "plain"),
    plot.title = element_text(size = 10, face = "plain", hjust = 0.5),
    legend.title = element_text(size = 10, face = "plain"),
    legend.text = element_text(size = 7),
    
    # Plot elements
    axis.line = element_line(size = 0.5, color = "black"),
    axis.ticks = element_line(size = 0.5, color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    
    # Legend positioning and appearance
    legend.position = "right",
    legend.box.background = element_rect(color = "white"),
    legend.background = element_rect(fill = "white", color = NA),
    
    # Panel appearance
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white")
  )

theme_set(publication_theme)
```

Analyse the TF enriched in most/least human selected sequences compared to the AL pool. 

This notebook take as input the `sea.tsv` file obtained by doing enrichment analysis with SEA with the HOCOMOCO 13 as TF motifs database and thresdholding on FDR 0.01. 

Filtering for K562 TF using the GTEX (V8) Whole Blood annotations. HOCOMOCO13 is using Uniprot IDs while GTEX uses gene symbols, the `tf_masterlist.tsv` file was downloaded from the Autosome.org website and contains the TF Uniprot IDs and their corresponding gene symbols.

# Data import and preprocessing

Selection of TFs expressed in K562

```{r}
sea.files= list.files("sequence_analysis/SEA", "sea.tsv", recursive = T, full.names = T)

#Select TF expressed in K562
gtex = read.table("reference/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz", header = T, sep = "\t",skip = 2)
gtex.selected.TF = subset(gtex, (Whole.Blood>=1))$Description

#Trim the table to remove the mouse orthologs 
tf.masterlist = read.table("reference/tf_masterlist.tsv", header = T, sep = "\t", fill = T)
tf.masterlist = tf.masterlist[grep("HUMAN", tf.masterlist$curated.uniprot_id),]
tf.masterlist$curated.uniprot_id = str_split(tf.masterlist$curated.uniprot_id, "_",simplify = T)[,1]
tf.masterlist = tf.masterlist[,c("curated.uniprot_id","auto.gene_symbol")]

#Select for TFs expressed in WB
#Merging with the TF masterlist from the autosome team is important because the TFs IDs are from uniprot and not gene symbols
gtex = merge(gtex[,c("Description","Whole.Blood")], tf.masterlist, by.x="Description", by.y="auto.gene_symbol")
gtex.selected.TF = subset(gtex, (Whole.Blood>=1))$curated.uniprot_id
```

```{r}
import.sea = function(sea.tsv){
  sea.df = read.table(sea.tsv, sep = "\t", header = T)
  #Format the TF names
  sea.df$ALT_ID =  sea.df$ID
  sea.df$ID = str_split(sea.df$ID, "\\.", simplify = T)[,1]
  sea.df = subset(sea.df, ID %in% gtex.selected.TF)
  
  sea.df = sea.df[order(sea.df$ENR_RATIO,decreasing = T),]
  sea.df = sea.df[!(duplicated(sea.df$ID)),]
  sea.df$ID = factor(sea.df$ID, levels = unique(sea.df$ID))
  sea.df$ALT_ID = factor(sea.df$ALT_ID, levels = unique(sea.df$ALT_ID))
  
  return(sea.df)
}
```

```{r}
sea.tnb = lapply(sea.files, function(x){import.sea(x)})
names(sea.tnb) = c("Least diverse (1x60)", "Most diverse (1x60)", "Least diverse (3x20)", "Most diverse (3x20)","Least uncertain (1x60)", "Most uncertain (1x60)", "Least uncertain (3x20)", "Most uncertain (3x20)")

rounds = rep(c("1x60","1x60","3x20","3x20"),2)
info = rep(c("Least","Most"),4)
al.method = c(rep("Diverse",4), rep("Uncertain",4))

for(i in seq_along(sea.tnb)){
  sea.tnb[[i]]$DB = names(sea.tnb)[i]
  sea.tnb[[i]]$rounds = rounds[i]
  sea.tnb[[i]]$info = info[i]
  sea.tnb[[i]]$al.method = al.method[i]
}

sea.tnb.df = do.call("rbind", sea.tnb)

sea.tnb.3x20 = sea.tnb.df[grep( "\\(3x20\\)",sea.tnb.df$DB),]
sea.tnb.3x20$DB = factor(sea.tnb.3x20$DB, levels = c("Least uncertain (3x20)", "Most uncertain (3x20)", "Least diverse (3x20)", "Most diverse (3x20)"))
sea.tnb.1x60 = sea.tnb.df[grep( "\\(1x60\\)",sea.tnb.df$DB),]
sea.tnb.1x60$DB = factor(sea.tnb.1x60$DB, levels = c("Least uncertain (1x60)", "Most uncertain (1x60)", "Least diverse (1x60)", "Most diverse (1x60)"))
```

# Jitter plots

```{r}
library(ggrepel)

top10.3x20 <-  do.call("rbind",lapply(group_split(sea.tnb.3x20, DB), function(x){
  df = x %>%
    filter(ENR_RATIO > 2) %>%
    arrange(desc(ENR_RATIO)) %>%
    head(10)
}))



jitter.enr.3x20 = ggplot(sea.tnb.3x20 , aes(x=DB, y=ENR_RATIO, colour=info))+
  geom_jitter(size=.5)+
  xlab("")+
  ylab("TFBS enrichment over the AL pool")+
  ggtitle("Human - 3x20")+
  geom_text_repel(data = top10.3x20, aes(x=DB, y=ENR_RATIO, label = ID), size=2.5, max.overlaps=20, show.legend=FALSE, colour="black")+
    theme(aspect.ratio = c(0.85/1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10),legend.position = "none")

top10.1x60 <-  do.call("rbind",lapply(group_split(sea.tnb.1x60, DB), function(x){
  df = x %>%
    filter(ENR_RATIO > 2) %>%
    arrange(desc(ENR_RATIO)) %>%
    head(10)
}))

jitter.enr.1x60 = ggplot(sea.tnb.1x60 , aes(x=DB, y=ENR_RATIO, colour=info))+
  geom_jitter(size=.5)+
  geom_text_repel(data = top10.1x60, aes(x=DB, y=ENR_RATIO, label = ID), size=2.5, max.overlaps=20, show.legend=FALSE, colour="black")+
  xlab("")+
  ylab("TFBS enrichment over the AL pool")+
  ggtitle("Human - 1x60")+
    theme(aspect.ratio = c(0.85/1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10),legend.position = "none")
jitter.enr.3x20
jitter.enr.1x60
```


