---
title: "AL selected yeast sequences expression analysis Report"
author: "Emmanuel Cazottes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,          # Show code chunks in output
  warning = FALSE,      # Don't show warnings
  message = FALSE,      # Don't show messages
  fig.width = 10,       # Default figure width
  fig.height = 6,       # Default figure height
  fig.align = "center"  # Center align figures
)

# Load commonly used packages
library(stringr)
library(tidyverse)
library(reshape2)
library(ggpubr)
library(ggsci)
library(gridExtra)

# Custom ggplot theme settings
publication_theme <- theme_pubr() +
  theme(
    # Text elements
    axis.text = element_text(size = 7, color = "black"),
    axis.title = element_text(size = 10, face = "plain"),
    plot.title = element_text(size = 10, face = "plain", hjust = 0.5),
    legend.title = element_text(size = 10, face = "plain"),
    legend.text = element_text(size = 7),
    
    # Plot elements
    axis.line = element_line(size = 0.5, color = "black"),
    axis.ticks = element_line(size = 0.5, color = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    
    # Legend positioning and appearance
    legend.position = "right",
    legend.box.background = element_rect(color = "white"),
    legend.background = element_rect(fill = "white", color = NA),
    
    # Panel appearance
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white")
  )

theme_set(publication_theme)
```

This notebook analyze the expression levels of the most and least selected yeast sequences by AC methods (top/bottom 1%) compared to random selection and the AL pool. Yeast GPRA data from Rafi et al 2024 are used to annotate expression of selectede sequences.

# Data import and preprocessing

Sequences expression as measured in yeast GRPA
```{r}
mpra = read.table("reference/GPRA_Rafi2024.txt", header = F, sep = "\t", col.names = c("Seq", "expression"))

mpra$IDs = seq(0, c(nrow(mpra)-1), 1) #Number the sequences to give them IDs
mpra = mpra[,-1]
```

Load sequences

```{r}
load.sequences = function(seq.tab){
  seq = read.table(seq.tab, header = F, sep = "\t", col.names = c("IDs", "sequence","method"))
  seq = seq[,-2]
  return(seq)
}

#Top and bottom sequences
tnb = load.sequences("sequences/yeast_top_n_bottom_1.trim.tsv")
tnb[grep("1x60", tnb$method),"rounds"] = "1x60k"
tnb[grep("1x60", tnb$method,invert = T),"rounds"] = "3x20k"
tnb[grep("uncertainty_((1x60)|(3x20))_bottom.*", tnb$method),"al.method"] = "Least uncertain"
tnb[grep("uncertainty_((1x60)|(3x20))_top.*", tnb$method),"al.method"] = "Most uncertain"
tnb[grep("diversity_((1x60)|(3x20))_bottom.*", tnb$method),"al.method"] = "Least diverse"
tnb[grep("diversity_((1x60)|(3x20))_top.*", tnb$method),"al.method"] = "Most diverse"

#AL pool
alPool = load.sequences("sequences/yeast_pool.trim.tsv")
alPool$rounds = "AL pool"
alPool$method = "AL pool"
alPool$al.method = "AL pool"
```

Merge the mpra data with the sequence data

```{r}
tnb.mpra = merge(tnb, mpra,by="IDs")
alPool.mpra = merge(alPool, mpra, by="IDs")

#Random selection
rnd.mpra = alPool.mpra[sample(seq(1,nrow(alPool.mpra),1),100000),]
rnd.mpra$method = "Random selection"
rnd.mpra$rounds = "Random selection"
rnd.mpra$al.method = "Random selection"

```

Bind everything together for plotting

```{r}
tnb.mpra.3x20 = subset(tnb.mpra, rounds=="3x20k")
tnb.mpra.1x60 = subset(tnb.mpra, rounds=="1x60k")
```


```{r}
seq.mpra.3x20 = rbind( tnb.mpra.3x20, rnd.mpra, alPool.mpra)
seq.mpra.1x60 = rbind(tnb.mpra.1x60, rnd.mpra, alPool.mpra)

#Levels for plotting
seq.mpra.3x20$al.method = factor(seq.mpra.3x20$al.method, levels = c("AL pool", "Random selection",  "Least uncertain","Most uncertain","Least diverse","Most diverse"))
seq.mpra.1x60$al.method = factor(seq.mpra.1x60$al.method, levels = c("AL pool", "Random selection",  "Least uncertain","Most uncertain","Least diverse","Most diverse"))
```

```{r}
# compute the Mann-Whitney pvalue
mannWhitneyTest = function(data){
  df.pval = map_dfr(levels(data$al.method), function(level){
  group1 = subset(data, al.method==level)$expression
  group2 = subset(data, al.method=="AL pool")$expression
  
  mwt = wilcox.test(group1, group2)
  
  data.frame(al.method=level,
             pvalue=signif(mwt$p.value, digits = 2),
             statistic=mwt$statistic)
  })
  
  df.pval$pvalue = ifelse((df.pval$pvalue==0 | df.pval$pvalue<=2.2e-16),"p < 2.2e-16", paste0("p = ",df.pval$pvalue)) #R reports 0 because the pvalue is smaller than R's standard floating-point precision

  return(df.pval)
  }
```

```{r}
pval60k = mannWhitneyTest(seq.mpra.1x60)
pval60k$y.axis = 18
pval3x20 = mannWhitneyTest(seq.mpra.3x20)
pval3x20$y.axis = 18
```

# Plots

```{r}
boxplot.fc.3x20 = ggplot(seq.mpra.3x20 , aes(x=al.method, y=expression))+
  geom_violin(width=1.25, adjust=3)+  
  geom_boxplot(width=0.1,outliers=F)+
    geom_text(data = pval3x20, aes(x=al.method,y=y.axis, label = pvalue),size = 3)+
  xlab("")+
  ylab("Expression")+
  ggtitle("Yeast - 3x20")+
    theme(aspect.ratio = c(0.85/1),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10))

boxplot.fc.1x60 = ggplot(seq.mpra.1x60 , aes(x=al.method, y=expression))+
  geom_violin(width=1.25, adjust=3)+  
  geom_boxplot(width=0.1,outliers=F)+
    geom_text(data = pval60k, aes(x=al.method,y=y.axis, label = pvalue),size = 3)+
  xlab("")+
  ylab("Expression")+
    ggtitle("Yeast - 1x60")+
    theme(aspect.ratio = c(0.85/1),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.margin = margin(5, 5, 5, 5, "mm"),
        axis.text = element_text(size=7), axis.title = element_text(size=10))

```
```{r selected-sequences-plots, fig.cap="Selected sequences expression analysis"}
print(boxplot.fc.3x20)
print(boxplot.fc.1x60)
```



